Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Multi-Notch Searches

### 2017-10-16 paper.tex
tsnslmns

It is useful to carefully describe the two commonly used database search approaches, prior to defining multi-notch searches and the impact that they have on G-PTM-D performance.
A traditional search in bottom-up proteomics considers potential matches between theoretical peptides and experimental spectra if the candidate peptide mass and the observed precursor mass are equal within some defined tolerance (e.g. 5ppm, or 2.1 Da).
We refer to this type of search here as a \textit{single-notch} search.
The other common search is a \textit{wide-parent tolerance search} (e.g. +/- 500Da)\citep{Chick_2015,Kong_2017,Li_2016}.
In searches of that type, the precursor mass and the mass of the best matching theoretical peptide could be many Da apart.
We describe an alternative to these search modes: the \textit{multi-notch} search.
This search is an extension of the single-notch search that allows a variety of multiple specific mass differences between precursor mass and theoretical mass (See Figure~\ref{fig:fig3-searchTypes}) where the mass difference is easily interpretable.
It can also be seen as a narrowing of the wide-parent tolerance search to only include prespecified mass differences.
The advantages and disadvantages of this alternative strategy are described below.

@@

The traditional single-notch search limits the range of acceptable mass differences between a candidate peptide and the observed experimental precursor in order to find an exact match to the observed peptide in a protein database.
A mass tolerance of 1-20ppm is a commonly used, resulting in a highly restrictive search.
It is not uncommon for the precursor mass tolerance of single-notch searches to be relaxed to as much as 2 or 3 Da.
One rationale for this is that errors in the deconvolution of MS1 spectra frequently result in an experimental precursor mass reported being different by 1.0029 or 2.0054 Da from the true isolated peptide mass.
The relaxed tolerance approach identifies those peptides correctly.
A second rationale for this approach is that multiple peptides are frequently co-isolated and co-fragmented with all peptide fragments being reported in the same tandem mass spectrum.
Frequently the only well-fragmented peptide is different from the one corresponding to the isolated peak reported by the instrument, and the relaxed tolerance approach may identify this peptide, provided the reported charge state matches.
The major limitation of this approach is that a 2 or 3 Da tolerance includes mass shifts that, when present, may indicate that the matched peptide is not the correct identification, arguably defeating the purpose of a narrow-mass search.
This may happen when the mass difference arises due to identified peptide being different from the actual isolated one by a low mass modification or an amino acid substitution.

The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4mouse-1012}, generated by compiling the results of a 2.5 Da tolerance single-notch search.

An alternative to these two traditional single-notch searches is a limited multi-notch search that allows mass differences between precursor and theoretical peptide of 0, 1.0029 or 2.0054 Da but limits the width about any one of those three tolerances to a few ppm (1-20 ppm).
This approach provides a means of getting an increased number of correct identifications but mitigates problems with wrong identifications.
This alternative would appear to result in loss of identification of co-isolated peptides, but MetaMorpheus is programmed to mitigate this problem by identifying multiple co-isolated precursor masses, see section below.

To qualitatively illustrate the described advantages, we conducted the three searches on all 28 calibrated Jurkat fractions.
The numbers of confidently identified PSMs and misassigned PSMs are reported in Table~\ref{tbl:singleVsMultiNotch}.

>>>>

It is useful to carefully describe the two commonly used database search approaches, prior to defining multi-notch searches and the impact that they have on G-PTM-D performance.
A traditional search in bottom-up proteomics considers potential matches between theoretical peptides and experimental spectra if the candidate peptide mass and the observed precursor mass are equal within some defined tolerance (e.g. 5ppm, or 2.1 Da).
We refer to this type of search here as a \textit{single-notch} search.
The other common search is a \textit{wide-parent tolerance search} (e.g. $\pm 500$ Da)\citep{Chick_2015,Kong_2017,Li_2016}.
In searches of that type, the precursor mass and the mass of the best matching theoretical peptide could be many Da apart.
We describe an alternative to these search modes: the \textit{multi-notch} search.
This search is an extension of the single-notch search that allows a variety of multiple specific mass differences between precursor mass and theoretical mass (See Figure~\ref{fig:fig3-searchTypes}) where the mass difference is easily interpretable.
It can also be seen as a narrowing of the wide-parent tolerance search to only include prespecified mass differences.
The advantages and disadvantages of this alternative strategy are described below.

@@

The traditional single-notch search limits the range of acceptable mass differences between a candidate peptide and the observed experimental precursor in order to find an exact match to the observed peptide in a protein database.
A mass tolerance of <20ppm is commonly used on high-resolution MS1 data, resulting in a highly restrictive search.
It is not uncommon for the precursor mass tolerance of single-notch searches to be relaxed to as much as 2 or 3 Da.
One rationale for this is that errors in the deisotoping of MS1 spectra frequently result in a reported experimental precursor mass being different by 1.0029 or 2.0054 Da from the true isolated peptide mass.
The relaxed tolerance approach identifies those peptides correctly.
A second rationale for this approach is that multiple peptides are frequently co-isolated and co-fragmented with all peptide fragments being observed in the same tandem mass spectrum.
Frequently, the best-fragmented peptide is different from the one corresponding to the precursor mass reported by the instrument, and the relaxed tolerance approach may identify this peptide, provided the reported charge state matches.
The major limitation of this approach is that a 2 or 3 Da tolerance includes mass shifts that, when present, may indicate that the matched peptide is not the correct identification, arguably defeating the purpose of a narrow-mass search.
This may happen when the mass difference arises due to the identified peptide being different from the actual isolated precursor by a low mass modification or amino acid substitutions.

The described issues are apparent in mass difference histograms in Figure~\ref{fig:fig4mouse-1012}, generated by compiling the results of a 2.5 Da tolerance single-notch search.
These plots clearly show that numerous peptides have missed monoisotopic (Mm) peaks, deamidation, or an amino acid substitution.

We propose a limited multi-notch search as a remedy to the trade-offs caused by choosing a narrow (few ppm) or wider (few Da) precursor mass tolerance during a traditional single-notch search.
This limited multi-notch search allows mass differences between precursor and theoretical peptides of 0, 1.0029 and 2.0054 Da, but it limits the width around these three mass differences to a few ppm (1-20 ppm).
This approach provides a means of getting an increased number of correct identifications while mitigating problems with incorrect identifications.
This alternative would appear to result in loss of identification of co-isolated peptides, but MetaMorpheus is programmed to handle this problem by identifying multiple co-isolated precursor masses (see section below).

To illustrate the described advantages, we compared the three types of searches using all 28 calibrated Jurkat fractions (Table~\ref{tbl:singleVsMultiNotch}).
The limited notch search yields the largest number of confidently identified PSMs with only a small fraction being misassinged.
We consider PSMs to be misassigned if the G-PTM-D procedure determines a higher-scoring modified peptide for a PSM, see the last section for details.
